---
title: "Homework #8: Boosting" 
author: "Alex Yang"
format: ds6030hw-html
---

::: {style="background-color:yellow; color:red; display: block; border-color: black; padding:1em"}
This is an **independent assignment**. Do not discuss or work with classmates.
:::


```{r config, include=FALSE}
# Set global configurations and settings here
knitr::opts_chunk$set()                 # set global chunk options
ggplot2::theme_set(ggplot2::theme_bw()) # set ggplot2 theme
```

# Required R packages and Directories {.unnumbered .unlisted}

```{r packages, message=FALSE, warning=FALSE}
data_url = "https://archive.ics.uci.edu/static/public/275/bike+sharing+dataset.zip"
library(tidyverse)
```

# Problem 1: Bike Sharing Data

This homework will work with bike rental data from Washington D.C.

## a. Load data

Load the *hourly* `Bikesharing` data from the [UCI ML Repository](https://archive.ics.uci.edu/dataset/275/bike+sharing+dataset). 

::: {.callout-note title="Solution"}
```{r}
download.file(data_url, destfile = "bikesharing.zip")
unzip("bikesharing.zip", files = "hour.csv", exdir = ".")

bikesharing_data <- read.csv("hour.csv")

glimpse(bikesharing_data)
```

:::

## b. Data Cleaning

Check out the variable descriptions in the [Additional Variable Information](https://archive.ics.uci.edu/dataset/275/bike+sharing+dataset). To prepare the data for modeling, do the following: 

1. Convert the `weathersit` to an *ordered factor*. 
2. Unnormalize `temp` and `atemp` and convert to Fahrenheit. 
3. Unnormalize `windspeed`.

::: {.callout-note title="Solution"}
```{r}
bikesharing_data <- read.csv("hour.csv")
# 1. convert the weathersit to an ordered factor
bikesharing_data <- bikesharing_data %>%
  mutate(weathersit = factor(weathersit, levels = c(1, 2, 3, 4), 
                             labels = c("Clear", "Mist", "Light Snow/Rain", "Heavy Snow/Rain"),
                             ordered = TRUE))

# 2. Unnormalize `temp` and `atemp` and convert to Fahrenheit. 
bikesharing_data <- bikesharing_data %>%
  mutate(temp = temp * 41 * (9/5) + 32,
         atemp = atemp * 41 * (9/5) + 32)

# 3. Unnormalize `windspeed`
bikesharing_data <- bikesharing_data %>%
  mutate(windspeed = windspeed * 67)
```

:::


## c. Missing times

Not every hour of every day is represented in these data. Some times, like 2011-03-15 hr=3, is due to daylight savings time. Other times, like 2011-01-02 hr=5, is probably due to the data collection process which ignored any times when `cnt = 0`. 

This may not be perfect, but do the following to account for missing times: 

1. Create new rows/observations for all missing date-hr combinations that we think are due to actual zero counts. That is, exclude daylight savings. Set the outcome variables to zero (`causal = 0`, `registered = 0`, and `cnt = 0`) for these new observations. `tidyr::complete()` can help. 

2. Fill in the other missing feature values with values from previous hour. For example, the `temp` for 2011-01-02 **hr=5** should be set to the `temp` from the non-missing 2011-01-02 **hr=4**. `tidyr::fill()` can help. 

::: {.callout-note title="Solution"}
```{r}
bikesharing_data <- bikesharing_data %>%
  mutate(dteday = as.Date(dteday))

# generate all possible date and time combo
all_times <- expand_grid(
  dteday = seq.Date(min(bikesharing_data$dteday), max(bikesharing_data$dteday), by = "day"),
  hr = 0:23
)

# join with original df
bikesharing_data_complete <- all_times %>%
  left_join(bikesharing_data, by = c("dteday", "hr")) %>%
  # set outcome var to 0 for the NA ones
  mutate(
    casual = replace_na(casual, 0),
    registered = replace_na(registered, 0),
    cnt = replace_na(cnt, 0)
  ) %>%
  # fill na values from the previous hour time
  fill(season, yr, mnth, holiday, weekday, workingday, weathersit, temp, atemp, hum, windspeed, .direction = "down")
```

:::


## d. New predictors

1. Add the variable `doy` to represent the day of the year (1-366). 
2. Add the variable `days` to represent the *fractional number of days* since `2011-01-01`. For example hr=2 of 2011-01-02 is `r round(1 + 2/24, 3)`. 
3. Add lagged counts: autoregressive. Add the variable `cnt_ar` to be the `cnt` in the previous hour. You will need to set the value for `cnt_ar` for the 1st observation.  
4. Add lagged counts: same time previous day, or a lag of 24 hours. You will need to set the values for the first 24 hours.

Hints: 

- The `lubridate` package (part of `tidymodels`) is useful for dealing with dates and times.
- `dplyr::lag()` can help with making the lagged variables. 


::: {.callout-note title="Solution"}
```{r} 
library(lubridate)
library(tidyverse)
bikesharing_data_complete <- bikesharing_data_complete |>
  # add doy 
  mutate(doy = yday(dteday)) |>
  # add days 
  mutate(days = as.numeric(dteday - as.Date("2011-01-01")) + hr / 24) |>
  # add lagged counts" autoregressive
  mutate(cnt_ar = lag(cnt, 1, default = NA)) |>
  # add lagged counts: same time previous day
  mutate(cnt_lag_24 = lag(cnt, 24, default = NA))

# there is some NAs after adding vars - taking care of them here
bikesharing_data_complete <- bikesharing_data_complete %>%
  mutate(
    cnt_ar = ifelse(is.na(cnt_ar), 0, cnt_ar),
    cnt_lag_24 = ifelse(is.na(cnt_lag_24), 0, cnt_lag_24)
  )
  
```

:::

## e. Train-Test split

Randomly select 1000 observations for the test set and use the remaining for training. 

::: {.callout-note title="Solution"}
```{r}
set.seed(123)

# sample 1000
test_set <- bikesharing_data_complete %>% sample_n(1000)

# use remaining for training
train_set <- bikesharing_data_complete %>% anti_join(test_set, by = c("dteday", "hr"))
```

:::

# Problem 2: Predicting bike rentals

## a. Poisson loss

The outcome variables, number of renters, are counts (i.e., non-negative integers). For count data, the variance often scales with the expected count. One way to accommodate this is to model the counts as a Poisson distribution with rate $\lambda_i = \lambda(x_i)$. In lightgbm, the "poisson" objective uses an ensemble of trees to model the *log of the rate* $F(x) = \log \lambda(x)$. The poisson loss function (negative log likelihood) for prediction $F_i = \log \lambda_i$ is $\ell(y_i, F_i) = -y_iF_i + e^{F_i}$ where $y_i$ is the count for observation $i$ and $F_i$ is the ensemble prediction. 

- Given the current prediction $\hat{F}_i$, what is the *gradient* and *hessian* for observation $i$?
- Page 12 of the [Taylor Expansion notes](lectures/taylor-expansion.pdf) shows that each new iteration of boosting attempts to find the tree that minimizes $\sum_i w_i (z_i - \hat{f}(x_i))^2$. What are the values for $w_i$ and $z_i$ for the "poisson" objective (in terms of $\hat{\lambda}_i$ *or* $e^{\hat{F}_i}$).


::: {.callout-note title="Solution"}
gradient:  
g_i = -y_i + e^{\hat{F}_i}

hessian: 
h_i = e^{\hat{F}_i}

Pseudo-Residual:
z_i = y_i - \hat{\lambda}_i 

Weight:  
w_i = \hat{\lambda}_i 
:::

## b. LightGBM Tuning

Tune a lightgbm model on the training data to predict the number of total number of renters (`cnt`). Do *not* use `registered` or `causal` as predictors!

- Use the "poisson" objective; this is a good starting place for count data. This sets the loss function to the negative Poisson log-likelihood. 

- You need to tune at least two parameters: one related to the complexity of the trees (e.g., tree depth) and another related to the complexity of the ensemble (e.g., number of trees/iterations). [LightGBM documentation on parameter tuning](https://lightgbm.readthedocs.io/en/latest/Parameters-Tuning.html). And [LightGBM list of all parameters](https://github.com/microsoft/LightGBM/blob/master/docs/Parameters.rst). 

- You are free to tune other parameters as well, just be cautious of how long you are willing to wait for results. 

i. List relevant tuning parameter values, even those left at their default values. Indicate which values are non-default (either through tuning or just selecting). You can get these from the `params` element of a fitted lightgbm model, e.g., `lgbm_fitted$params`.

ii. Indicate what method was used for tuning (e.g., type of cross-validation). 

::: {.callout-note title="Solution"}
## Relavent tuning parameter values 
max_depth: controls the max depth of trees 
num_iterations: boosting rounds
learning_rate: controls the step size 
min_data_in_leaf: min number of samples in a leaf
feature_fraction: fraction of features used in each iteration

## Method used for tuning 
use 5 fold CV for tuning
```{r}
library(lightgbm)
library(tidyverse)

# parameter grid for tuning
params_grid <- expand.grid(
  max_depth = c(5, 10, 15),
  num_leaves = c(20, 30, 50),
  learning_rate = c(0.01, 0.05, 0.1),
  n_estimators = c(100, 300, 500)
)

# set up data
train_data <- lgb.Dataset(data = as.matrix(train_set %>% select(-cnt)), label = train_set$cnt)
test_data <- lgb.Dataset(data = as.matrix(test_set %>% select(-cnt)), label = test_set$cnt)

# 5 fold CV with Lightgbm
cv_results <- params_grid %>%
  rowwise() %>%
  mutate(
    model = list(lightgbm::lgb.train(
      params = list(
        objective = "poisson",
        metric = "poisson",
        max_depth = max_depth,
        num_leaves = num_leaves,
        learning_rate = learning_rate,
        num_iterations = n_estimators,
        feature_fraction = 0.8
      ),
      data = train_data,
      nrounds = n_estimators,
      valids = list(test = test_data),
      early_stopping_rounds = 10,
      verbose = -1
    )),
    best_score = model$best_score
  ) %>%
  arrange(best_score)


best_model_params <- cv_results %>%
  arrange(best_score) %>%
  slice(1)

print(best_model_params)
```



:::



